Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-17 background.tex
+

\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables2.png}
    \vspace{-2.5em}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
    \vspace{-1em}
\end{figure}

@@

\paragraph{\add{Mapping a New Page}}
\add{Mapping a new page into a virtual address space is a key piece fo functionality, and also a key challenge for verification.
This operation takes a chunk of physical memory to install, and a (page-aligned) virtual address at which to make it
visible to the process.
Starting from the page table root (\lstinline|cr3| in Figure \ref{fig:pagetables}), it must traverse the page tables until it locates
the correct location of the L1 entry which should point to the physical page (installing new L3, L2, and L1 tables as needed),
and update the L1 entry to point to the physical page. This is complicated by the fact that the kernel accesses the page
tables through \emph{virtual} addresses, but the page table root and the addresses in the L4--L1 entries are all \emph{physical}
addresses. So the mapping process must repeatedly perform \emph{backwards} mappings (not supported by hardware) from
physical addresses of each level's entry to a virtual address that is then mapped back to the corresponding physical address!
And for modularity, the translation process (and possible allocation of missing tables) is typically implemented in a function
used for every level of translation (i.e., a level-polymorphic function).
These compounding subtleties are one reason \textsc{CertiKOS} and \textsc{seL4} \emph{trust} their page table
manipulations; both the reasoning to prove correctness, and the kernel invariants that ensure correctness of typical implementations,
are remarkably subtle.
}
\looseness=-1

@@

We suppress many details of \iris assertions for brevity, but briefly note
a few recurring details used heavily in \iris but not necessarily in traditional
separation logics.
Because \iris is an embedded framework in \rocq, proofs in \iris-derived
logics (including ours) often encapsulate raw \rocq assertions: $\ulcorner P \urcorner$ is an embedding
of the \rocq assertion $P$ into an \iris assertion (used for things like equality and other
general pure logical assertions), similar to prior \rocq-embedded program logics~\cite{Chlipala2013Bedrock}. Newly-defined \iris assertions are in fact
\rocq terms of a particular type, rather than being drawn from a fixed vocabulary.
\del{We introduce other details as they arise.}
\iris includes two forms of implication. The magic wand operator $\wand$ is an affine implication:
$A\wand B$ describes a resource which, if combined with a resource satisfying $A$, will satisfy $B$.
Notably, this implication involves no changes to ghost state. \iris, building on the Views framework~\cite{Dinsdale-Young2013Views},
also includes a \emph{view shift} operator $\vs$ which models updates to ghost state: $A \vs B$
means resources satisfying $A$ may be transformed into resources satisfying $B$, intuitively by updating only ghost state.\footnote{\iris
experts may note that this is \emph{technically} mildly misleading given how \iris's update modalities and weakest precondition
are defined, but
this is adequate intuition for non-experts in \iris to follow their use.
}
View shifts are essentially logical entailment plus ghost updates.
\looseness=-1

@@

An important limitation \del{here} is that to date, every separation logic
has assumed that all pointer addresses are meant for use in a single address space, which avoided the
problem of tracking that certain points-to assertions are true only for certain address spaces, but not others.

@@

Of particular interest for reasoning about virtual memory are modalities that permit \emph{naming} the alternate 
circumstances, prominently \emph{hybrid} modal logics~\cite{blackburn1995hybrid,areces2001hybrid}, which come equipped 
with assertions of the form $[\ell](P)$ indicating that $P$ is true in the specific alternate circumstance (Kripke world)
 named by the \emph{nominal} $\ell$. Note that a distinctive property of hybrid logics is that, rather than hiding
the points at which a modal assertion is evaluated inside the modality's definition, the choice of what world a modalized
assertion should be true in is chosen \emph{in the assertion itself}. This allows assertions to talk about not simply whether some other assertion
is true in some possible future or past world related in a fixed way to the current world, but to talk about \emph{arbitrary}
other worlds.
\add{This is somewhat different from the accessibility relation modalities more common in verification, but
also well-established.}
This explicit naming of alternate worlds increases the power of modal logics \add{over those without it}~\cite{blackburn1995hybrid}, and is actually
necessary for completeness in classical separation logics~\cite{brotherston2014parametric}.
%   A key distinguishing feature of
% hybrid logic is that the logic itself does not fix the set of modalities, but is defined relative to any set of sensible worlds.

@@

\add{This is essential to updating page mappings for the current address space in use by a CPU.}

>>>>

\begin{figure}
%    \begin{framed}
%\column{.59\textwidth}    
\begin{subfigure}

\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-0.5,xscale=0.5]
%uncomment if require: \path (0,430); %set diagram left start at 0, and has height of 430

%Shape: Rectangle [id:dp9406932326447903] 
\draw   (33,141) -- (123,141) -- (123,292) -- (33,292) -- cycle ;
%Shape: Rectangle [id:dp7682116051394419] 
\draw   (166,141) -- (256,141) -- (256,292) -- (166,292) -- cycle ;
%Shape: Rectangle [id:dp07376733516930423] 
\draw   (295,141) -- (385,141) -- (385,292) -- (295,292) -- cycle ;
%Shape: Rectangle [id:dp7106629147891306] 
\draw   (427,141) -- (517,141) -- (517,292) -- (427,292) -- cycle ;
%Shape: Rectangle [id:dp1342906826203598] 
\draw   (115,54) -- (185,54) -- (185,94) -- (115,94) -- cycle ;
%Shape: Rectangle [id:dp08072439007207399] 
\draw   (185,54) -- (255,54) -- (255,94) -- (185,94) -- cycle ;
%Shape: Rectangle [id:dp8282933603833216] 
\draw   (255,54) -- (325,54) -- (325,94) -- (255,94) -- cycle ;
%Shape: Rectangle [id:dp8416255090865326] 
\draw   (325,54) -- (395,54) -- (395,94) -- (325,94) -- cycle ;
%Shape: Rectangle [id:dp22911958218529804] 
\draw   (395,54) -- (465,54) -- (465,94) -- (395,94) -- cycle ;
%Shape: Rectangle [id:dp33732955213990756] 
\draw   (465,54) -- (535,54) -- (535,94) -- (465,94) -- cycle ;
%Curve Lines [id:da6688730321284677] 
\draw    (211.5,96) .. controls (-36.44,98.94) and (15.75,190.24) .. (30.17,213.65) ;
\draw [shift={(31,215)}, rotate = 238.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Shape: Rectangle [id:dp6044747978986409] 
\draw   (33,191) -- (123,191) -- (123,231) -- (33,231) -- cycle ;
%Curve Lines [id:da8867562998134584] 
\draw    (289.5,95) .. controls (115.15,110.76) and (137.77,175.03) .. (163.33,197.98) ;
\draw [shift={(164.5,199)}, rotate = 220.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Shape: Rectangle [id:dp8861153167245819] 
\draw   (166,177) -- (256,177) -- (256,217) -- (166,217) -- cycle ;
%Shape: Rectangle [id:dp05979844200848894] 
\draw  [dash pattern={on 4.5pt off 4.5pt}] (295,156) -- (385,156) -- (385,196) -- (295,196) -- cycle ;
%Shape: Rectangle [id:dp17403733244247843] 
\draw   (427,212) -- (517,212) -- (517,252) -- (427,252) -- cycle ;
%Curve Lines [id:da20581124900573955] 
\draw [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ] [dash pattern={on 4.5pt off 4.5pt}]  (357,97) .. controls (262.92,110.72) and (272.56,148.45) .. (291.81,173.48) ;
\draw [shift={(293,175)}, rotate = 231.34] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da07589611603635316] 
\draw    (425.5,95) .. controls (380.42,119.5) and (406.41,202.58) .. (422.53,224.72) ;
\draw [shift={(423.5,226)}, rotate = 231.34] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da6852167546673456] 
\draw    (497.5,95) .. controls (564.48,91.06) and (565,230.71) .. (567.87,269.31) ;
\draw [shift={(568,271)}, rotate = 265.24] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da017521077423665377] 
\draw    (39.5,359) .. controls (-12.71,369.84) and (17.07,323.43) .. (32.31,293.36) ;
\draw [shift={(33,292)}, rotate = 116.57] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da6750042432035159] 
\draw    (122,215) .. controls (128.9,254.4) and (131.91,269.54) .. (159.71,289.1) ;
\draw [shift={(161,290)}, rotate = 214.59] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da14788158682113428] 
\draw    (257,197) .. controls (258.95,236) and (277.06,270.25) .. (293.72,290.47) ;
\draw [shift={(295,292)}, rotate = 229.64] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da3297977595767969] 
\draw    (386,177) .. controls (392.86,226) and (408.36,270.2) .. (425.92,290.77) ;
\draw [shift={(427,292)}, rotate = 228.01] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Curve Lines [id:da7710391961792686] 
\draw    (517,229) .. controls (524.76,253.25) and (540.05,269.03) .. (552.82,282.73) ;
\draw [shift={(554,284)}, rotate = 227.12] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%Shape: Rectangle [id:dp1535611125971872] 
\draw   (514,336) -- (584,336) -- (584,376) -- (514,376) -- cycle ;
%Shape: Rectangle [id:dp05427947476226991] 
\draw   (95,338) -- (136.5,338) -- (136.5,378) -- (95,378) -- cycle ;
%Shape: Rectangle [id:dp10511138743752002] 
\draw   (136.5,338) -- (265.5,338) -- (265.5,378) -- (136.5,378) -- cycle ;
%Shape: Rectangle [id:dp8917680919534263] 
\draw   (265.5,338) -- (305,338) -- (305,378) -- (265.5,378) -- cycle ;
\draw   (558,284.5) .. controls (558,279.25) and (562.25,275) .. (567.5,275) .. controls (572.75,275) and (577,279.25) .. (577,284.5) .. controls (577,289.75) and (572.75,294) .. (567.5,294) .. controls (562.25,294) and (558,289.75) .. (558,284.5) -- cycle ; \draw   (558,284.5) -- (577,284.5) ; \draw   (567.5,275) -- (567.5,294) ;
%Curve Lines [id:da9312068388962929] 
\draw    (568,298) .. controls (584.66,309.76) and (586.91,318.64) .. (565.35,335.93) ;
\draw [shift={(564,337)}, rotate = 321.95] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;

% Text Node
\draw (11,14) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Virtual}};
% Text Node
\draw (570,395) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Physical}};
% Text Node
\draw (104,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 63}};
% Text Node
\draw (175,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 48}};
% Text Node
\draw (245,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 39}};
% Text Node
\draw (314,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 30}};
% Text Node
\draw (385,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 21}};
% Text Node
\draw (454,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 12}};
% Text Node
\draw (523,28) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 0}};
% Text Node
\draw (132,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Sign}};
% Text Node
\draw (195,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L4 Offset}};
% Text Node
\draw (264,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L3 Offset}};
% Text Node
\draw (336,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L2 Offset}};
% Text Node
\draw (406,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L1 Offset}};
% Text Node
\draw (472,65) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Page Offset}};
% Text Node
\draw (67,63) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize va}};
% Text Node
\draw (53,203) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L4 Entry}};
% Text Node
\draw (185,189) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L3 Entry}};
% Text Node
\draw (316,167) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L2 Entry}};
% Text Node
\draw (449,223) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize L1 Entry}};
% Text Node
\draw (51,347) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize root}};
% Text Node
\draw (465,349) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize page}};
% Text Node
\draw (532,346) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 4K Data}};
% Text Node
\draw (156,386) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 51}};
% Text Node
\draw (226,386) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize 12}};
% Text Node
\draw (145,350) node [anchor=north west][inner sep=0.75pt]   [align=left] {{\scriptsize Base Addr}};

\end{tikzpicture}
          \caption{Accessing to the Page Referenced by L1 Entry}
        \label{fig:enter-label}
    \end{subfigure}
    \vfill
    \begin{multicols}{2}
%    \begin{figure}
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
/*
@param entry: entry whose frame address is used to locate the a-level-lower entry in the page-tree
@return next: return a virtual pointer to the descendant tree
*/
pte_t *pte_get_next_table(pte_t *entry) {
  pte_t *next;
  /*Reading page table entry justified by virtual pte-pointsto*/
  if (!entry->present(*\label{line:read_entry_contentsC}*) (*\label{line:conditional_childrenC}*)) {
    /* Jump if the present bit is not zero; no need to allocate next level (* \label{line:check_entry_presentC} *)(* \label{line:mask_presentC} *)  (*\label{line:after_concluding_qfrac1C}*)*/
    /*Allocation path starts here (*\label{line:pass_addrof_nextC}*)(*\label{line:alloc_path_startC}*)*/
    pte_initialize(entry);(* \label{line:call_to_pte_initializeC} *)
    /*Entry value updates:*/
    entry->pfn = nextpaddr;
    entry->present = 1; (*\label{line:install_new_entryC}*)
    /*Now we know that entry is initialized, so we satisfy the condition to access children list */
  } (*\label{line:end_of_allocation_pathCB}*)
  /*Physical to virtual address mapping code segment starts here (*\label{line:finalpieceSB}*)*/
  uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn)
  uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr); (*\label{line:p2vCB}*);
  next = (pte_t *) next_virt_addr;(*\label{line:finalpieceEB}*)
  return next;
}
\end{lstlisting}
\columnbreak
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
pte_t *walkpgdir(pte_t *l4, void *va){ 
 pte_t *l4_entry = &l4[L4Offset(va)];
 pte_t *l3 = pte_nxt_table(l4_entry);
 pte_t *l3_entry = &l3[L3Offset(va)];
 pte_t *l2 = pte_nxt_table(l3_entry);
 pte_t *l2_entry = &l2[L2Offset(va)];  
 pte_t *l1 = pte_nxt_table(l2_entry);
 pte_t *l1_entry = &l1[L1Offset(va)];  
 return l1_entry
}
\end{lstlisting}
%          \caption{Walking page-table tree to locate L1 entry C-source code.}
%        \label{fig:enter-label}
 %   \end{figure}
\end{multicols}
%\end{framed}
    \vspace{-2.5em}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
    \vspace{-1em}
\end{figure}

@@

@@

We suppress many details of \iris assertions for brevity, but briefly note
a few recurring details used heavily in \iris but not necessarily in traditional
separation logics.
Because \iris is an embedded framework in \rocq, proofs in \iris-derived
logics (including ours) often encapsulate raw \rocq assertions: $\ulcorner P \urcorner$ is an embedding
of the \rocq assertion $P$ into an \iris assertion (used for things like equality and other
general pure logical assertions), similar to prior \rocq-embedded program logics~\cite{Chlipala2013Bedrock}. Newly-defined \iris assertions are in fact
\rocq terms of a particular type, rather than being drawn from a fixed vocabulary.
We introduce other details as they arise.

\iris includes two forms of implication. The magic wand operator $\wand$ is an affine implication:
$A\wand B$ describes a resource which, if combined with a resource satisfying $A$, will satisfy $B$.
Notably, this implication involves no changes to ghost state. \iris, building on the Views framework~\cite{Dinsdale-Young2013Views},
also includes a \emph{view shift} operator $\vs$ which models updates to ghost state: $A \vs B$
means resources satisfying $A$ may be transformed into resources satisfying $B$, intuitively by updating only ghost state.\footnote{\iris
experts may note that this is \emph{technically} mildly misleading given how \iris's update modalities and weakest precondition
are defined, but
this is adequate intuition for non-experts in \iris to follow their use.
}
\looseness=-1

@@

An important limitation here is that to date, every separation logic
has assumed that all pointer addresses are meant for use in a single address space, which avoided the
problem of tracking that certain points-to assertions are true only for certain address spaces, but not others.

@@

Of particular interest for reasoning about virtual memory are modalities that permit \emph{naming} the alternate 
circumstances, prominently \emph{hybrid} modal logics~\cite{blackburn1995hybrid,areces2001hybrid}, which come equipped 
with assertions of the form $[\ell](P)$ indicating that $P$ is true in the specific alternate circumstance (Kripke world)
 named by the \emph{nominal} $\ell$. Note that a distinctive property of hybrid logics is that, rather than hiding
the points at which a modal assertion is evaluated inside the modality's definition, the choice of what world a modalized
assertion should be true in is chosen \emph{in the assertion itself}. This allows assertions to talk about not simply whether some other assertion
is true in some possible future or past world related in a fixed way to the current world, but to talk about \emph{arbitrary}
other worlds. This explicit naming of alternate worlds increases the power of modal logics~\cite{blackburn1995hybrid}, and is actually
necessary for completeness in classical separation logics~\cite{brotherston2014parametric}.
%   A key distinguishing feature of
% hybrid logic is that the logic itself does not fix the set of modalities, but is defined relative to any set of sensible worlds.

@@

### 2025-07-17 examples_C.tex
+

\lstdefinestyle{mystyle}
{
    language = C,
    basicstyle = {\footnotesize,\color{black}},
    backgroundcolor = {\color{white}},
    stringstyle = {\color{string-color}},
    keywordstyle = {\color{key-color}},
    keywordstyle = [2]{\color{lime}},
    keywordstyle = [3]{\color{yellow}},
    keywordstyle = [4]{\color{teal}},
    otherkeywords = {;,<<,>>,++},
    morekeywords = [2]{;},
    morekeywords = [3]{<<, >>},
    morekeywords = [4]{++},
    escapeinside={(*}{*)},
}
\newcommand{\sumwalkabsent}{
  \ownGhost\gammaPred{\authfrag{\singletonMap{\texttt{entry+KERNBASE}}{(\textsf{qfrac}, \textsf{entry})}}}
}

>>>>

\newcommand{\sumwalkabsent}{
  \ownGhost\gammaPred{\authfrag{\singletonMap{\texttt{entry+KERNBASE}}{(\textsf{qfrac}, \textsf{entry})}}}
}
